---
title: "Computational Statistics & Probability"
subtitle: "Problem Set 2 - Linear Models"
author: "Author: Neelesh Bhalla  \nCollaborators: Nils Marthiensen, Chia-Jung Chang"
date: "`r Sys.Date()`"
output: 
  pdf_document:
    latex_engine: xelatex
---

## 1. Multiple Regression & Causal Models
Return to the Howell1 dataset and consider the causal relationship between age and weight in children. Let’s define children as anyone younger than 13 and assume that age influences weight directly and through age-related physical changes that occur during development – physical attributes that a child’s height will serve as proxy. We may summarize this causal background knowledge by the DAG:
```{r}
library(rethinking)
library(dagitty)
dag_q1 <- dagitty('dag{ W <- A -> H -> W }')
drawdag( dag_q1 )
```
where Ai is age of child i, Hi is height of child i and W is weight of child i.

a) What is the total causal effect of year-by-year growth of !Kung children on their weight? Construct a linear regression (m1a) to estimate the total causal effect of each year of growth on a !Kung child’s weight. Assume average birth weight is 4kg. Use prior predictive simulation to assess the implications of your priors.
```{r}
impliedConditionalIndependencies(dag_q1)
#There are no conditional independence, so there is no output to display
```

```{r}
data(Howell1)
d <- Howell1
children <- d[ d$age < 13 , ]
plot( children$weight ~ children$age )
# taking an idea of the relation between age and weight 
```

```{r}
N <- 100
a <- rnorm( N , 12 , 4 )
b1 <- rlnorm( N , 0 , 1 )
```

```{r}
# Prior predictive simulation for the weight and age model
plot( NULL , xlim=range(children$age) , ylim=c(-50,50) ,
xlab="age" , ylab="weight" )
abline( h=4 , lty=2, lwd=0.5 )
abline( h=35 , lty=2 , lwd=0.5 )
mtext( "log(b1) ~ dnorm(0,1)" )
xbar <- mean(children$age)
for ( i in 1:N ) curve( a[i] + b1[i]*(x - xbar) ,
from=min(children$age) , to=max(children$age) , add=TRUE ,
col=col.alpha("black",0.2) )
```


```{r}
# Posterior distribution for weight and age model
xbar <- mean(children$age)
# fit model
m1a <- quap(
alist(
weight ~ dnorm( mu , sigma ) ,
mu <- a + exp(log_b1)*( age - xbar ) ,
a ~ dnorm( 12 , 4 ) ,
log_b1 ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=children )
```

```{r}
precis( m1a )
# the marginal posterior distributions
```

```{r}
round( vcov( m1a ) , 3 )
# the variance-covariance matrix
```

```{r}
plot( weight ~ age , data=children , col=rangi2 )
post <- extract.samples( m1a )
a_map <- mean(post$a)
b1_map <- exp(mean(post$log_b1))
curve( a_map + b1_map*(x - xbar) , add=TRUE )
```

b) What is the total causal effect of height on weight? Construct a linear regression (m1b) to estimate the total causal effect height on a !Kung child’s weight. Use prior predictive simulation to assess the implication of your priors.
```{r}
plot( children$weight ~ children$height )
# taking an idea of the relation between height and weight 
```

```{r}
N <- 100
a <- rnorm( N , 12 , 4)
b2 <- rlnorm( N , 0 , 1 )
```

```{r}
# Prior predictive simulation for the weight and height model
plot( NULL , xlim=range(children$height) , ylim=c(-50,50) ,
xlab="height" , ylab="weight" )
abline( h=0 , lty=2, lwd=0.5 )
abline( h=35 , lty=2 , lwd=0.5 )
mtext( "log(b2) ~ dnorm(0,1)" )
xbar <- mean(children$height)
for ( i in 1:N ) curve( a[i] + b2[i]*(x - xbar) ,
from=min(children$height) , to=max(children$height) , add=TRUE ,
col=col.alpha("black",0.2) )
```


```{r}
# Posterior distribution for weight and height model
xbar <- mean(children$height)
# fit model
m1b <- quap(
alist(
weight ~ dnorm( mu , sigma ) ,
mu <- a + exp(log_b2)*( height - xbar ) ,
a ~ dnorm( 12 , 4 ) ,
log_b2 ~ dnorm( 0 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) ,
data=children )
```

```{r}
precis( m1b )
# the marginal posterior distributions
```

```{r}
round( vcov( m1b ) , 3 )
# the variance-covariance matrix
```

```{r}
plot( weight ~ height , data=children , col=rangi2 )
post <- extract.samples( m1b )
a_map <- mean(post$a)
b2_map <- exp(mean(post$log_b2))
curve( a_map + b2_map*(x - xbar) , add=TRUE )
```

c) After knowing the age of a !Kung child, what additional value is there in also knowing the child’s height? Conversely, after knowing the height of a !Kung child, what additional value is there in also knowing the child’s age?
```{r}
children$A <- scale( children$age )
children$H <- scale( children$height )
children$W <- scale( children$weight )
#fit models
m1_age <- quap(
alist(
W ~ dnorm( mu , sigma ) ,
mu <- a + bA * A ,
a ~ dnorm( 0 , 0.2 ) ,
bA ~ dnorm( 0 , 0.5 ) ,
sigma ~ dexp( 1 )
) , data = children )
m1_height <- quap(
alist(
W ~ dnorm( mu , sigma ) ,
mu <- a + bH * H ,
a ~ dnorm( 0 , 0.2 ) ,
bH ~ dnorm( 0 , 0.5 ) ,
sigma ~ dexp( 1 )
) , data = children )
m1_all <- quap(
alist(
W ~ dnorm( mu , sigma ) ,
mu <- a + bA*A + bH*H ,
a ~ dnorm( 0 , 0.2 ) ,
bA ~ dnorm( 0 , 0.5 ) ,
bH ~ dnorm( 0 , 0.5 ) ,
sigma ~ dexp( 1 )
) , data = children )
# Visualize results
```

```{r}
precis( m1_all )
```

```{r}
round( vcov( m1_all ) , 3 )
# the variance-covariance matrix
```

```{r}
library(arulesViz)
library(raster)
library(Rgraphviz)
plot( coeftab(m1_age,m1_height,m1_all), par=c("bA","bH") )
```
```{r}
# from the above plot it is evident that height has a bigger stronger 
# overall influence on the weight of the children. Thus the causal relationship 
# is stronger between height-weight
# Thus, one can conclude that after knowing the height of a !Kung child, 
# there is almost little additional value in also knowing the child’s age?
```


## 2. Causal Influence with Categorical Variables
The causal relationship between age and weight might be different for girls and boys.

a) To investigate whether this is so, construct a single linear regression with a categorical variable for sex to estimate the total causal effect of age on weight separately for !Kung boys and girls. Plot your data and overlay the two regression lines, one for girls and one for boys.
```{r}
data(Howell1)
d <- Howell1
children <- d[ d$age < 13 , ]
```

```{r}
children$sex <- ifelse( children$male==1 , 2 , 1 )
str( children$sex )
# indexing --> now “1” means female and “2” means male
```
```{r}
xbar <- mean(children$age)
# fit model
m2a <- quap(
alist(
weight ~ dnorm( mu , sigma ) ,
mu <- a[sex] + b[sex]*( age - xbar ) ,
a[sex] ~ dnorm( 4 , 1 ) ,
b[sex] ~ dlnorm( 0.5 , 1 ) ,
sigma ~ dunif( 0 , 10 )
) , data=children )
# here both 'a' and 'b' are assumed to be dependent on the sex
# It is explainable because the weights of a boy child and a girl child
# are different at birth. Moreover, the rate of growth of weights are also 
# different for both genders.
```

```{r}
# plot data and both regression lines
plot( weight ~ age , data=children , col=c("red","blue")[children$sex])
post <- extract.samples( m2a )
a_map_girl <- mean(post$a[,1])
a_map_boy <- mean(post$a[,2])
b_map_girl <- mean(post$b[,1])
b_map_boy <- mean(post$b[,2])
curve( a_map_girl + b_map_girl*(x - xbar) , add=TRUE , col="pink")
curve( a_map_boy + b_map_boy*(x - xbar) , add=TRUE , col="blue")
legend(x = "topleft",
legend = c("girl", "boy"),
lty = c(1, 1),
col = c("pink", "blue"),
lwd = 2)
```

```{r}
precis( m2a , depth=2 )
# the marginal posterior distributions
```

```{r}
round( vcov( m2a ) , 3 )
# the variance-covariance matrix
```

b) Do they differ? If so, provide one or more posterior contrasts as a summary.
```{r}
seq <- 0:12
mu1 <- sim(m2a,data=list(age=seq,sex=rep(1,13)))
mu2 <- sim(m2a,data=list(age=seq,sex=rep(2,13)))
mu_contrast <- mu1
for ( i in 1:13 ) mu_contrast[,i] <- mu2[,i] - mu1[,i]
plot( x=seq, y=(colMeans(mu2)-colMeans(mu1)), type="l" , xlim=c(0,13) , ylim=c(-15,15) , xlab="age" ,
ylab="weight difference (boys-girls)" )
for ( p in c(0.5,0.67,0.89,0.99) ) # credibility intervals
shade( apply(mu_contrast,2,PI,prob=p) , seq )
abline(h=0,lty=2,lwd=2)
```


