---
title: "Computational Statistics & Probability"
author: "Lab 1 - Bayesian Inference"
date: "Fall 2022"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
answer_key <- TRUE
library(rethinking)
```

## 1. Posterior Probability Distributions

Use the following R code to generate a set of `samples` to from which to answer questions about
its distribution.

```{r posterior prob}
p_grid <-seq( from=0 , to=1, length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6, size=9, prob=p_grid )
posterior <-likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(212)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace = TRUE)
```

**a)**  How much posterior probability lies below $p = 0.25$?

```{r posterior prob a, include=answer_key}
sum(samples < 0.25) / length(samples)
```
 

**b)**  How much posterior probability lies above $p = 0.75$?

```{r posterior prob b, include=answer_key}
sum(samples > 0.75) / length(samples)
```

**c)** How much posterior probability lies between $p = 0.25$ and $p = 0.75$?

```{r posterior prob c, include=answer_key}
sum(samples > 0.25 & samples < 0.75) / length(samples)
```

**d)** 25% of the posterior probability lies below which value of $p$?

```{r posterior prob d, include=answer_key}
quantile(samples, 0.25)
```


## 2. Globe Tossing Experiment

**a)**  Suppose the globe tossing experiment yielded the following sequence of 15 observations,

$$ [ W, L, W, W, L, L, W, L, W, L, L, W, L, W, W ] $$
where $W$ denotes 'water' and $L$ denotes 'land'. Using grid approximation and the same flat prior as in Question 1, construct the posterior distribution.


```{r globe toss a, include=answer_key}
p_grid <-seq( from=0 , to=1, length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 8, size=15, prob=p_grid )
posterior <-likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(212)
samples_a <- sample( p_grid , prob=posterior , size=1e4 , replace = TRUE)
```

**b)**  Using grid approximation, construct the posterior distribution with a prior that is 0 below $p =0.5$ and otherwise constant.

```{r globe toss b, include=answer_key}
p_grid <-seq( from=0 , to=1, length.out=1000 )
prior <- ifelse(p_grid < 0.5, 0, 1) 
likelihood <- dbinom( 8, size=15, prob=p_grid )
posterior <-likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(212)
samples_b <- sample( p_grid , prob=posterior , size=1e4 , replace = TRUE)
```


**c)**  Explain the difference between model (2a) and (2b).

```{r globe toss c, include=answer_key}
# Model (b) encodes your prior knowledge that at least one-half of the Earth's 
# surface is covered with water, whereas Model (a) encodes that you believe any 
# proportion, from all water to all land, is equally plausible
```

**d)** Which prior, (a) or (b), is better and why?

```{r globe toss d, include=answer_key}
# The more informative prior, (b), gives an estimate that is closer to the true
# value (~ 71%):
mean(samples_a)
mean(samples_b)

# Nevertheless, Model (b) still concentrates more mass between 0.5 and 0.6 than
# we would like. Observing more data would allow either model to converge to the
# true value of $p \approx 0.71$

```


```{r globe toss PLOT, include=answer_key}
# PLOT
dens(samples_a, xlab="probability p", xlim=c(0,1), ylim=c(0,5.6), col="blue3")
dens( samples_b, add=TRUE, col="red3")
abline( v=0.71, lty=1, col=grau(0.5),lwd=2)
text(0.78, 4, "p = 0.71", col=grau(0.7))
text(0.3, 1.5, "sample_a", col="blue3")
text(0.42, 4, "sample_b", col="red3")

```

